pirouette: the error BEAST2 makes in inferring a phylogeny

### 2019-02-28 article.tex
Add example 4 with figure

\usepackage{listings}
\usepackage{hyperref}
\usepackage{todonotes}
\usepackage{verbatim}
\usepackage{pgf}
\usepackage{bm}

@@

Since BEAST performs a bayesian analysis, it needs not only genetic data but also tree priors, describing how the diversification occur, to return posterior phylogenies.
BEAST works using standard tree priors like Yule or birth-death \giovanni{integrate this if necessary}.
The recent development of BEAST2 [\cite{bouckaert2014beast}] provides, unlike its predecessor, the possibility for third-party users to integrate its prior libraries with new ones. Many of these have been developed, in the form of likelihood diversification models: birth-death models that have rates that are 
constant [\cite{nee1994reconstructed}], 
time-dependent [\cite{nee1994reconstructed}, \cite{rabosky2008explosive}], 
diversity-dependent [\cite{etienne2011diversity}],
or can shift [the SLS model. Laudanno et al., in preparation].
There are birth death models in which speciation is a process that 
takes time [\cite{rosindell2010protracted}][\cite{etienne2012prolonging}],
can co-occur [Laudanno et al., in preparation],
dependent on a trait that is binary [\cite{maddison2007estimating}], 
has multiple states [\cite{fitzjohn2012diversitree}],
has a concealed state [\cite{beaulieu2016detecting}] 
or has multiple observed and concealed states [\cite{herrera2018detecting}].

@@

From a given phylogeny, the likelihood is maximized
to infer the best estimates for the parameter values.
For such  models, it is a standard practice to assess their performance on a controlled environment: this is done by simulating phylogenies from known parameters, then comparing the estimations obtained by maximum likelihood with the original parameters.

\richel{
  We should bridge this method with how pirouette does things differently;
  why the maximum likelihood does not apply in this context  
}
\giovanni{
I will try writing a first draft of this
}
 If the likelihood proves to be effective to describe the process, it can be implemented into BEAST2 as a new prior. Despite that it is still possible that already existing BEAST2 priors will perform well enough and, therefore, that the introduction of a new model is not necessary.

@@

\verb;pirouette; is an R package that performs such an experiment,
and is built on \verb;babette; [Bilderbeek \& Etienne, 2018], 
which calls BEAST2 [\cite{bouckaert2014beast}]. 
With \verb;pirouette;, one
can easily measure the error made by Bayesian inference in recovering
any given phylogeny, helping us to evaluate the necessity of a new diversification model.
Once verified that the introduction of a new model is indeed beneficial,
\iffalse
Upon finding a diversification model is warranted, 
\fi
it can then be added to BEAST2's existing set of (birth-death and 
coalescent) models via the BEAST2 package system.

@@

The first step simulates a DNA alignment from the given phylogeny.
One can specify a DNA sequence
of any length at the root of the phylogeny, a DNA mutation rate, a
site (i.e. nucleotide substitution) model, 
a clock model, a random number generator (RNG) seed and a location
where the alignment is saved to. This step is relatively fast, but longer
DNA alignments will noticeably slow down the inference step.

@@

The third step infers a Bayesian posterior from the simulated alignment,
using the inference model(s) selected in the previous step. The user
can specify the additional parameters needed for the BEAST2 run, which
are the Markov-Chain Monte Carlo (MCMC) setup, 
an optional MRCA prior and an RNG seed.
The MCMC setup determines the number of posterior trees sampled.
An MRCA prior allows the inferred phylogeny to have a dated crown age.

The fourth step measures the inference error, using the phylogenies in the
Bayesian posterior. These phylogenies are compared to the given
original phylogeny using an error statistic, which is the nLTT 
statistic (\cite{janzen2015approximate}) by default,
but also a custom error statistic is supplied, based
on the absolute difference in the gamma statistic [\cite{pybus2000testing}]. 
Additionally, the user can specify the
proportion of posterior phylogenies to 
discard (i.e. the burn-in), throwing away the first $10\%$
of all phylogenies by default. This burn-in is used to discard
the part of the MCMC chain that has not yet converged to a
representative part of the state space.

@@

\iffalse
An optional step is to generate a 'twin tree', that will be
analyzed in the same way as the true tree.
The twin tree has the same topology as the given phylogeny, yet its branch lengths follow the same distribution as a Yule (pure-birth) or (constant rate) birth-death tree prior 
\giovanni{let's revise this after we meet Rampal and define how a twin tree should be created}.
The use of such a twin tree, is to assess the minimum level of noise (i.e. error) BEAST2 makes when the input (the twin phylogeny) has an ideal branch length distribution.
\fi

An optional step is to generate a 'twin tree', that will be
analyzed in the same way as the true tree.
The twin tree has the same topology as the given phylogeny, yet its branching times are simulated according to the tree prior used by BEAST2 to produce a posterior.
The goal of this procedure is to generate a tree to use as a control for the main experiment.
The procedure to generate a twin tree is the following:

From the given phylogeny we maximize a birth-death likelihood [\cite{nee1994reconstructed}] to infer the standard birth-death rates, namely speciation rate $\lambda$ and extinction rate $\mu$.

We use $\lambda$ and $\mu$ to simulate a new birth-death tree, conditioned on having the same number of tips as the original tree. We store its branching times.

We then combine such branching times with the original tree's topology to obtain the twin tree.

@@

\begin{lstlisting}[
    language=R, floatplacement=H, frame=single, 
    label = {lst:create_ideal_tree}, 
    caption = {Create an ideal tree}
  ]
phylogeny <- create_ideal_tree(n_taxa = 5, crown_age = 10)
\end{lstlisting}

\begin{figure}[h]
  \includegraphics[width=\textwidth]{figure_bd.png}
  \caption{The ideal tree, as created by listing \ref{lst:create_ideal_tree}}
\end{figure}

@@

\begin{figure}[h]
  \includegraphics[width=\textwidth]{figure_example_1.png}
  \caption{
    The error BEAST2 makes from a phylogeny 
    when the generative and inference model are the same.
  }
  \label{fig:example_1}
\end{figure}

@@

\giovanni{After having explained that we are going to use as candidates all the possible inference models, it comes natural to wonder what is the best inference model selected by marginal likelihoods. It could be nice maybe, adding the name in the lengend on top-right, in place of the generic "candidate" word. It could be also nice to add it in the caption and/or in the main text.}

\begin{figure}[h]
  \includegraphics[width=\textwidth]{figure_example_2.png}
  \caption{
    The error BEAST2 makes from a phylogeny when
    picking the best inference model
  }
  \label{fig:example_2}
\end{figure}

@@

\begin{figure}[h]
  \includegraphics[width=\textwidth]{figure_example_3.png}
  \caption{
    The error BEAST2 makes from a phylogeny picking the best inference model versus the control.
    Here, the twin column shows the error BEAST2 makes starting from the original tree versus the twin. 
  }
  \label{fig:example_3}
\end{figure}

@@

@@

Here we discuss how to interpret the results of \verb;pirouette;.
Figure \ref{fig:example_1} shows that the generative model 
gives an error distribution with a highest density 
\giovanni{probably "median" is a better word}
at 0.02 
\giovanni{I would increase the precision to the second significative digit, e.g. 0.017},
where the range of errors (in this case, the nLTT statistic) is from 0.0 to 1.0 
\giovanni{providing standard deviation could be a better choice}.
The range of errors goes from 0.0 to 0.27 
\giovanni{this is contrast with the last sentence}. 
As we know the generative model, this error distribution is the minimum noise that BEAST2 produces for this generative model.
\richel{Show effective sample size}

Figure \ref{fig:example_2} \giovanni{same as RQ1} shows that the best candidate model gives an error distribution with a highest density at 0.02, where the range of errors (in this case, the nLTT statistic) is from 0.0 to 1.0. The range of errors goes from 0.0 to 0.08. 
\richel{Describe and show evidences for the candidate model}
\richel{Describe and show the best candidate model}
This means that it is rational to use this inference model to 
infer a posterior phylogeny distribution, 
as is done by \verb;pirouette;.
\richel{Show effective sample size}
\richel{Show posterior phylogenies as densitree}

Figure \ref{fig:example_3} \giovanni{same as RQ1} shows that the hand-picked model gives an error distribution with a highest density at 0.018, with a range from 0.0 to 0.1.
The twin tree has an error distribution with a highest density at 0.019,
with a range from 0.0 to 0.11, which is the noise caused by the tree prior. \giovanni{I am still wondering if "noise" is the right term to use here. I still don't have a precise opinion on that. The problem is that for me it's a bit hard to interpret this without a distribution...}
As there is little difference between the true and twin tree,
the (unknown) diversification model has little impact on the error
BEAST2 makes, hinting that the constant-rate birth-death model is
sufficiently good enough to explain the true phylogeny. 
The diversification model apparently has minor impact and need not to be
added to BEAST2.

\giovanni{since you refer, for every RQ, to the same values, it could be better to show them as a table}

>>>>

\usepackage{listings}
\usepackage{hyperref}
\usepackage{todonotes}
\usepackage{verbatim}
\usepackage{pgf}
\usepackage{bm}
\usepackage{multirow}

@@

Since BEAST performs a Bayesian analysis, 
it needs not only genetic data but also tree priors, 
describing how the diversification occur, to return posterior phylogenies.
BEAST works using standard tree priors like Yule or birth-death 
\giovanni{integrate this if necessary}.
The recent development of BEAST2 [\cite{bouckaert2014beast}] provides, 
unlike its predecessor, the possibility for third-party users 
to integrate its prior libraries with new ones. 
Many of these have been developed, 
in the form of likelihood diversification models: 
birth-death models that have rates that are 
constant [\cite{nee1994reconstructed}], 
time-dependent [\cite{nee1994reconstructed}, \cite{rabosky2008explosive}], 
diversity-dependent [\cite{etienne2011diversity}],
or can shift [the SLS model. Laudanno et al., in preparation].
There are birth death models in which speciation is a process that 
takes time [\cite{rosindell2010protracted}][\cite{etienne2012prolonging}],
can co-occur [Laudanno et al., in preparation],
dependent on a trait that is binary [\cite{maddison2007estimating}], 
has multiple states [\cite{fitzjohn2012diversitree}],
has a concealed state [\cite{beaulieu2016detecting}] 
or has multiple observed and concealed states [\cite{herrera2018detecting}].

@@

From a given phylogeny, the likelihood is maximized
to infer the best estimates for the parameter values.
For such  models, it is a standard practice to assess their performance 
on a controlled environment: this is done by simulating phylogenies from known parameters, 
then comparing the estimations obtained by maximum likelihood with the original parameters.
If the likelihood proves to be effective to describe the process, 
it can be implemented into BEAST2 as a new prior. 
Despite that it is still possible that already existing BEAST2 priors 
will perform well enough and, therefore, 
that the introduction of a new model is not necessary.

@@

\verb;pirouette; is an R package that performs such an experiment,
and is built on \verb;babette; [Bilderbeek \& Etienne, 2018], 
which calls BEAST2 [\cite{bouckaert2014beast}]. 
With \verb;pirouette;, one
can easily measure the error made by Bayesian inference in recovering
any given phylogeny, helping us to evaluate the necessity of a new diversification model.
Once verified that the introduction of a new model is indeed beneficial,
it can then be added to BEAST2's existing set of (birth-death and 
coalescent) models via the BEAST2 package system.

@@

The first step simulates a DNA alignment from the given 
phylogeny (Fig.~\ref{fig:pipeline} 1a $\rightarrow$ 2a).
One can specify a DNA sequence
of any length at the root of the phylogeny, a DNA mutation rate, a
site (i.e. nucleotide substitution) model, 
a clock model, a random number generator (RNG) seed and a location
where the alignment is saved to. This step is relatively fast, but longer
DNA alignments will noticeably slow down the inference step.

@@

The third step infers a Bayesian posterior from the simulated 
alignment (Fig.~\ref{fig:pipeline} 2a $\rightarrow$ 3a),
using the inference model(s) selected in the previous step. The user
can specify the additional parameters needed for the BEAST2 run, which
are the Markov-Chain Monte Carlo (MCMC) setup, 
an optional MRCA prior and an RNG seed.
The MCMC setup determines the number of posterior trees sampled.
An MRCA prior allows the inferred phylogeny to have a dated crown age.

The fourth step compares the true tree to each posterior tree
error using
the nLTT statistic (\cite{janzen2015approximate}) 
(Fig.~\ref{fig:pipeline} 3a $\rightarrow$ 4a), resulting
in an error distribution (Fig.~\ref{fig:pipeline} 4a $\rightarrow$ 5a).
\richel{I would not mind to remove 4a and 4b from the figure}
The nLTT statistic is used by default,
but also a user-defined error statistic can be used. As an example,
\verb;pirouette; suplies such a custom error statistic,
that uses an absolute difference in the gamma statistic [\cite{pybus2000testing}].
Additionally, the user can specify the
proportion of posterior phylogenies to 
discard (i.e. the burn-in), throwing away the first $10\%$
of all phylogenies by default. This burn-in is used to discard
the part of the MCMC chain that has not yet converged to a
representative part of the state space.

@@

An optional step is to generate a 'twin tree' (Fig.~\ref{fig:pipeline} 1a $\rightarrow$ 1b), 
that will be analyzed in the same way as the true tree.
The twin tree has the same topology as the given phylogeny, 
yet its branching times are simulated according to 
the tree prior used by BEAST2 to produce a posterior.
The goal of this procedure is to generate a tree 
to use as a control for the main experiment.
The procedure to generate a twin tree is the following:

From the given (true) phylogeny, we maximize a birth-death 
likelihood [\cite{nee1994reconstructed}] to infer the standard 
birth-death rates, namely speciation rate $\lambda$ and extinction rate $\mu$.
We use $\lambda$ and $\mu$ to simulate a new birth-death tree, 
conditioned on having the same number of tips as the original tree. 
We store its branching times. 
We then combine such branching times with the original tree's 
topology to obtain the twin tree.

@@

\begin{lstlisting}[
    language=R, floatplacement=H, frame=single, 
    label = {lst:create_yule_tree}, 
    caption = {Create a Yule tree}
  ]
phylogeny <- create_ideal_tree(n_taxa = 5, crown_age = 10)
\end{lstlisting}

\begin{figure}[h]
  \includegraphics[width=\textwidth]{figure_bd.png}
  \caption{The Yule tree, as created by listing \ref{lst:create_yule_tree}}
\end{figure}

@@

\begin{figure}[h]
  \includegraphics[width=\textwidth]{figure_example_1.png}
  \caption{
    Example 1: the error BEAST2 makes from a phylogeny 
    when the generative and inference model are the same.
    The x-axis specifies the tree type (which can be 'true' or 'twin'),
    the y-axis shows the error density
  }
  \label{fig:example_1}
\end{figure}

@@

\giovanni{describe winner}

\begin{figure}[h]
  \includegraphics[width=\textwidth]{figure_example_2.png}
  \caption{
    Example 2: the error BEAST2 makes from a phylogeny when
    picking the best inference model.
    The x-axis specifies the tree type (which can be 'true' or 'twin'),
    the y-axis shows the error density
  }
  \label{fig:example_2}
\end{figure}

@@

\begin{figure}[h]
  \includegraphics[width=\textwidth]{figure_example_3.png}
  \caption{
    Example 3: the error BEAST2 makes from a phylogeny picking the best inference model versus the control.
    Here, the twin column shows the error BEAST2 makes starting from the original tree versus the twin. 
    The x-axis specifies the tree type (which can be 'true' or 'twin'),
    the y-axis shows the error density
  }
  \label{fig:example_3}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Usage: fourth research question}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

A fourth research question that \verb;pirouette; answers, is:
"What is the error BEAST2 makes from a phylogeny for a known
generative model, compared to candidates with similar inference models,
in relationship to the background error?"

Here, same as example 2, we start with a tree generated by the Yule process,
by using the code shown in listing \ref{lst:create_yule_tree}.
The first step in \verb;pirouette; is to simulate a DNA alignment 
from the given phylogeny, for which we'll use the same code 
as shown in listing \ref{lst:create_alignment}.

\begin{table}
  \begin{tabular}{ | c | c | c | l | }
    \hline
    \textbf{model type} & \textbf{run if} & \textbf{measure evidence} & \textbf{inference model} \\ 
    \hline
    generative & always         & yes & JC, strict, Yule \\
    candidate  & best candidate & yes & JC, strict, BD \\
    candidate  & best candidate & yes & JC, strict, CCP \\
    candidate  & best candidate & yes & JC, strict, CEP \\
    \hline
  \end{tabular}
  \caption{
    Experimental setup to answer the fourth research question.
    JC: Jukes-Cantor site model.
    strict: strict clock model.
    Yule: Yule (pure-birth) tree prior.
    BD: birth-death tree prior.
    CCP: coalescent constant-population tree prior.
    CEP: coalescent exponential-population tree prior.
  }
  \label{tab:experiment_4}
\end{table}

In the second step we state our experiments. 
Within this example, we must both specify the generative and candidate 
experiments. Here we'll specify the generative model:

\begin{lstlisting}[language=R, floatplacement=H, frame=single]
generative_experiment <- create_experiment(
  model_type = "generative",
  run_if = "always",
  do_measure_evidence = TRUE
)
\end{lstlisting}

Specifying all candidates can be done in a factorial fashion using 
\verb;create_all_experiments;. In this case, all candidates will use the
same site and clock model, and only differ in their tree priors:

\begin{lstlisting}[language=R, floatplacement=H, frame=single]
candidate_experiments <- create_all_experiments(
  site_models = list(create_jc69_site_model()),
  clock_models = list(create_strict_clock_model()),
  tree_priors = list(
    create_bd_tree_prior(), 
    create_ccp_tree_prior(), 
    create_cep_tree_prior()
  )
)
\end{lstlisting}

We need to combine these experiments into one set:

\begin{lstlisting}[language=R, floatplacement=H, frame=single]
experiments <- c(list(generative_experiment), candidate_experiments)
\end{lstlisting}

Also here, the third (the BEAST2 inference) and fourth (measuring the error)
steps have sensible defaults. To get an idea of a baseline error made by
BEAST2, we also make use of the twinning option:

\begin{lstlisting}[language=R, floatplacement=H, frame=single]
pir_params <- create_pir_params(
  alignment_params = alignment_params,
  experiments = experiments,
  twinning_params = create_twinning_params()
)
\end{lstlisting}

The resulting figure is shown in figure \ref{fig:example_4}.

\giovanni{describe winner}

\begin{figure}[h]
  \includegraphics[width=\textwidth]{figure_example_4.png}
  \caption{
    Example 4: the error BEAST2 makes from a phylogeny for a known
    generative model, compared to candidates with similar inference models,
    in relationship to the background error.
    The x-axis specifies the tree type,
    the y-axis shows the error density
  }
  \label{fig:example_4}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}[h]
\centering
\begin{tabular}{ | r | l | l | l | l | }
\hline
\multirow{2}{*}{\textbf{Example}} & \multicolumn{2}{c|}{\textbf{True tree}} 
                                  & \multicolumn{2}{c|}{\textbf{Twin tree}} \\
\cline{2-5}
                                  & \textbf{HPD} & \textbf{range} & \textbf{HPD} & \textbf{range} \\
\hline
1  & $0.20$  & $0.0 - 0.27$  &         &               \\
2  & $0.02$  & $0.0 - 0.08$  &         &               \\
3  & $0.018$ & $0.0 - 0.1$   & $0.019$ & $0.0 - 0.11$  \\
4C & $0.22$  & $0.0 - 0.088$ & $0.009$ & $0.0 - 0.056$ \\
4G & $0.020$ & $0.0 - 0.075$ & $0.013$ & $0.0 - 0.052$ \\
\hline
\end{tabular}
\caption{Results of the examples. Errors are shows with the 
  highest posterior density first, and the range of the errors in
  round brackets second. Experiment 4 is split in two rows
  by either the candidate ('C') or generative model ('G') 
}
\label{tab:results}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Here we discuss how to interpret the results of \verb;pirouette;.
Figure \ref{fig:example_1} shows that the generative model 
gives an error distribution with a highest posterior density (HPD) 
\giovanni{probably "median" is a better word}
\richel{HPC is a valid Bayesian term I learned from the BEAST book}
at 0.02
\giovanni{I would increase the precision to the second significative digit, e.g. 0.017},
The range of errors goes from 0.0 to 0.27 
As we know the generative model, 
this error distribution is the minimum noise that BEAST2 produces 
for that generative model.
\richel{Show effective sample size}

Figure \ref{fig:example_2} shows that the best candidate model gives an error distribution 
with an HPD at 0.02. The range of errors goes from 0.0 to 0.08. 
\richel{Describe and show evidences for the candidate model}
\richel{Describe and show the best candidate model}
This means that it is rational to use this inference model to 
infer a posterior phylogeny distribution, 
as is done by \verb;pirouette;.
\richel{Show effective sample size}
\richel{Show posterior phylogenies as densitree}

Figure \ref{fig:example_3} shows that the hand-picked model 
gives an error distribution with an HPD at 0.018, 
with a range from 0.0 to 0.1.
The twin tree has an error distribution with a highest density at 0.019,
with a range from 0.0 to 0.11, which is an indication of the
error made by BEAST on that tree prior.
As there is little difference between the true and twin tree,
the (unknown) diversification model has little impact on the error
BEAST2 makes, hinting that the constant-rate birth-death model is
sufficiently good enough to explain the true phylogeny. 
The diversification model apparently has minor impact and need not to be
added to BEAST2.

\giovanni{
  since you refer, for every RQ, to the same values, 
  it could be better to show them as a table
}
